Retrofitting Effect Handlers onto OCaml

### 2020-11-11 retro-concurrency.tex
Updates

\newcommand{\rot}[2][75]{\adjustbox{angle=#1}{\textbf{#2}}}

@@

Our work is also timely. WebAssembly~\cite{Hass17} community group is actively
considering effect handlers as one of the mechanism for supporting concurrency,
asynchronous I/O and generators~\ref{WasmPropoal}. Project Loom~\cite{loom} is an
OpenJDK project which adds virtual threads and delimited continuations to Java.
Swift concurrency roadmap~\cite{swift} includes direct-style asynchronous
programming and structured concurrency as milestones. We believe that our
design choices will inform similar choices to be made in the other
industrial-strength languages.

@@

\begin{itemize}
  \item the design and implementation of one-shot effect handlers for OCaml.
    Our design retains OCaml's compatibility with program analysis tools that
    inspect the stack using DWARF unwind tables. We have validated our DWARF
    unwind tables using an automated validator tool~\cite{Bastian19}.
  \item a formal operational semantics for the effect handler implementation in
    OCaml.
  \item extensive evaluation which shows that our implementation has negligible
    impact on code that does not use effect handlers, and serves as an
    efficient foundation for scalable concurrent and parallel programming.
\end{itemize}

@@

The main challenge in implementing effect handlers in Multicore OCaml is
managing the program stack. To this end, in this section, we provide an
overview of the program stack and related mechanisms in stock OCaml.

@@

OCaml does not use the C calling convention. In particular, there are no
callee-saved registers in OCaml. On the |x86-64| backend, the C callee-saved
registers, |r15| holds the \emph{allocation pointer} into the minor heap used
for bump pointer allocation, and |r14| holds the |Caml_state|, a table of
global variables used by the runtime. This makes external calls extremely fast
in OCaml. If the external function does not allocate in the OCaml heap, then
the external functions can be called directly, and no bookkeeping is necessary.
For external functions which allocate in the OCaml heap, the cached allocation
pointer is saved to a field in the |Caml_state| before the external call, and
loaded back on return. Similarly, callbacks into OCaml are also cheap.
Callbacks into OCaml involves loading the arguments in the right registers and
calling the OCaml function. OCaml callbacks are relatively common as the
garbage collector (GC) finalisation functions are executed as OCaml callbacks.

@@

\begin{figure*}
  \begin{minipage}[t]{0.49\linewidth}
  \begin{smathpar}
  \begin{array}{rrcl}
    \text{Constants} & n & \coloneqq & ℤ \\
    \text{Abstractions} & \Lambda & \coloneqq & \lambda^o \mid \lambda^c \\
    \text{Expressions} & e & \coloneqq & n \mid x \mid e ~e \mid  \lam{x}{e} \mid e \odot e \mid \throw{l}{e} \\
                       &   & \mid      & \handle{e}{h} \mid \perform{l}{e} \\
    \text{Handlers} & h & \coloneqq & \{\caseval{x}{e}\} \mid \{\caseexn{l}{x}{e}\} \uplus h \\
                   &   & \mid & \{\caseeff{l}{x}{k}{e}\} \uplus h \\
    \text{Values} & v & \coloneqq & n \mid k \mid \clos{x}{e}{\env} \mid \effval{l}{k} \mid \exnval{l} \\
    \text{Frames} & r & \coloneqq & \farg{e}{\env} \mid \ffun{v} \mid \faritha{\odot}{e}{\env} \mid \farithb{\odot}{ℕ} \\
    \text{Environments} & \env & \coloneqq & \emptyset \mid \envext{\env}{x}{v} \\
  \end{array}
  \end{smathpar}
  \end{minipage}
  \begin{minipage}[t]{0.49\linewidth}
  \begin{smathpar}
  \begin{array}{rrcl}
    \text{Handler Closures} & \hc & \coloneqq & (h,\env) \\
    \text{Frame List} & \fl & \coloneqq & [] \mid r :: \fl \\
    \text{Fibers} & \fiber & \coloneqq & (\fl, \hc) \\
    \text{Continuations} & k & \coloneqq & [] \mid \fiber \kcons k \\
    \text{C stacks} & \cstack & \coloneqq & \cstacka{\fl}{\ostack}\\
    \text{OCaml stacks} & \ostack & \coloneqq & \ostacka{k}{\cstack} \mid \ostackemp \\
    \text{Stacks} & \stack & \coloneqq & \cstack \mid \ostack \\
    \text{Terms} & \term & \coloneqq & e \mid v \\
    \text{Configurations} & \config & \coloneqq & \configa{\tau}{\env}{\stack}
  \end{array}
  \end{smathpar}
  \end{minipage}
  \mycaption{Syntax of expressions and configurations}
  \label{sem:syntax}
\end{figure*}

@@

The operational semantics is an extension of a CEK machine
semantics~\cite{Felleisen86} for effect handlers, following the abstract machine
semantics of Hillerstrom et al.~\cite{Hillerstrom20}. The crucial difference
from Hillerstrom et al. is that our stacks are composed of alternating sequence
of OCaml and C stack segments. The program state is captured as configuration
$\config \coloneqq \configa{\tau}{\env}{\stack}$ with the current term $\tau$
under evaluation, its environment $\env$ and the current stack $\stack$. The
term is either an expression $e$ or a value $v$. The value are integer
constants $n$, continuations $k$, closure $\clos{x}{e}{\env}$, effect being
performed $\effval{l}{k}$ and exception being raised $\exnval{l}$. The
environment is a map from variables to values.

@@

The initial configuration for an expression $e$ is
$\configa{e}{\emptyset}{\cstacka{[]}{\bullet}}$, where the environment and the
stack are empty. The top-level reductions
\begin{smathpar}
	\begin{array}{cc}
		\textsc{StepC} ~
		\inferrule{(\term, \env, \fl, \ostack) \cstep \config}
							{\configa{\term}{\env}{\cstacka{\fl}{\ostack}} \step \config} &
		\textsc{StepO} ~
		\inferrule{(\term, \env, k, \cstack) \ostep \config}
							{\configa{\term}{\env}{\ostacka{k}{\cstack}} \step \config}
	\end{array}
\end{smathpar}
can be performed by either by taking a C step $\cstep$ or an OCaml step
$\ostep$. Intuitively, this splits the world into executing C code and OCaml
code.

\begin{figure*}
  \begin{smathpar}
    \begin{array}{rrcl}
      \textsc{Var}     & (x,\env, \fl) & \rightsquigarrow
                       & (\env(x), \env, \fl) \\
      \textsc{Arith1}  & (e_1 \odot e_2, \env, \fl) & \rightsquigarrow
                       & (e_1,\env, \faritha{\odot}{e_2}{\env}::\fl) \\
      \textsc{Arith2}  & (n_1, \_, \faritha{\odot}{e_2}{\env}::\fl) & \rightsquigarrow
                       & (e_2, \env, \farithb{\odot}{n_1}::\fl) \\
      \textsc{Arith3}  & (n_2, \env, \farithb{\odot}{n_1}::\fl) & \rightsquigarrow
                       & (\llbracket n_1 \odot n_2 \rrbracket, \env, \fl) \\
      \textsc{App1}    & (e_1 ~e_2, \env, \fl) & \rightsquigarrow
                       & (e_1, \env, \farg{e_2}{\env}::\fl) \\
      \textsc{App2}    & (\lam{x}{e}, \env, \fl) & \rightsquigarrow
                       & (\clos{x}{e}{\env}, \env, \fl) \\
      \textsc{App3}    & (\clos{x}{e_1}{\env_1}, \_, \farg{e_2}{\env_2}::\fl) & \rightsquigarrow
                       & (e_2, \env_2, \ffun{\clos{x}{e_1}{\env_1}}::\fl) \\
      \textsc{Resume1} & (k,\_,\farg{e_1}{\env_1}::\farg{e_2}{\env_2}::\fl) & \rightsquigarrow
                       & (e_1, \env_1, \ffun{k}::\farg{e_2}{\env_2}::\fl) \\
      \textsc{Resume2} & (\clos{x}{e_1}{\env_1}, \_, \ffun{k}::\farg{e_2}{\env_2}::\fl) & \rightsquigarrow
                       & (e_2, \env_2, \ffun{k}::\ffun{\clos{x}{e_1}{\env_1}}::\fl) \\
      \textsc{Perform} & (\perform{l}{e}, \env, \fl) & \rightsquigarrow
                       & (e, \env, \ffun{\effval{l}{[[],(\{\caseval{x}{x}\},\emptyset)]}}::\fl) \\
      \textsc{Raise}   & (\throw{l}{e}, \env, \fl) & \rightsquigarrow
                       & (e, \env, \ffun{\exnval{l}}::\fl)
    \end{array}
  \end{smathpar}
  \vspace{-1mm}
  \mycaption{Administrative Reductions -- $(\term, \env, \fl) \rightsquigarrow (\tau, \env, \fl)$.}
  \label{sem:step}
\end{figure*}
\begin{figure*}
  \begin{smathpar}
    \begin{array}{rrcll}
      \textsc{AdminC}   & (\term,\env,\fl,\ostack) & \cstep
                        & \configa{\term'}{\env'}{\cstacka{\fl'}{\ostack}}
                        & \text{if } (\term,\env,\fl) \rightsquigarrow (\term',\env',\fl') \\
      \textsc{RetToO}   & (v,\env,[],\ostacka{k}{\cstack}) & \cstep
                        & \configa{v}{\env}{\ostacka{k}{\cstack}} \\
      \textsc{CallC}    & (v, \_, \cclos{x}{e}{\env}::\fl, \ostack) & \cstep
                        & \configa{e}{\envext{\env}{x}{v}}{\cstacka{\fl}{\ostack}} \\
      \textsc{Callback} & (v, \_, \oclos{x}{e}{\env}::\fl, \ostack) & \cstep
                        & \configa{e}{\envext{\env}{x}{v}}{\ostacka{k}{\cstacka{\fl}{\ostack}}}
                        & \text{if } k = [[],(\{\caseval{x}{x}\},\emptyset)] \\
      \textsc{ExnFwdO}  & (v, \env, \ffun{\exnval{l}}::\_, \ostacka{(\fl,\hc) \kcons k}{\cstack}) & \cstep
                        & \configa{v}{\env}{\ostacka{(\ffun{\exnval{l}}::\fl,\hc) \kcons k}{\cstack}}
    \end{array}
  \end{smathpar}
  \vspace{-1mm}
  \mycaption{C Reductions -- $(\term, \env, \fl, \ostack) \cstep \config$.}
  \label{sem:cstep}
\end{figure*}

@@

\subsection{Heap-allocated fibers}

@@

When we first enter OCaml from C, a new fiber is allocated for the main OCaml
stack. Since callbacks may be frequent in OCaml programs that use finalisers,
we run the callbacks on the same fiber as the current one. For example, the
layout of the Multicore OCaml stack at |caml_raise_exn| in the |meander|
example from \S\ref{sec:stack} is shown in Figure~\ref{fig:mcstack}. The
functions |caml_call_c| and |caml_call_ocaml| switch the stacks, and hence are
shown in both the system stack and the fiber. Since we are reusing the fiber
for the callback, care must be taken to save and restore the |handler_info|
before calling and after returning from |c_to_ocaml| function, respectively.
The details of the exception frames have been elided in the figure. Thanks to
the fiber representation, external calls and callbacks remain fast.

\subsection{Effect handlers}
\label{sec:effimpl}

@@

\begin{table}
\caption{Micro benchmarks with handlers but no perform. Each entry is the
	slowdown factor ($\times$ times) over its idiomatic implementation in stock OCaml.}
\vspace{-3mm}
{
\begin{tabular}{r c c c c c}
	& \textbf{ack} & \textbf{fib} & \textbf{motzkin} & \textbf{sudan} & \textbf{tak} \\ \hline
	\texttt{MC} 	 	& 10.57 & 15.76 & 16.96 & 12.88 & 13.64 \\
	\texttt{monad} 	& 42.79 & 69.77 & 348.69 & 33.29 & 39.24 \\ \hline
\end{tabular}
}
\label{tab:micro_noperform}
	\vspace{-3mm}
\end{table}

@@

Kawahara et al.~\cite{Kawahara20} implement one-shot effect handlers using
coroutines (a non-local control flow mechanism) as a macro-expressible
translation, and present an embedding in Lua and Ruby. Lua provides asymmetric
coroutines~\cite{Lua} where each coroutine uses its own stack similar to how
each handled computation runs in its own fiber in Multicore OCaml. Similar to
Multicore OCaml, Lua does not allow coroutines to include C frames. It is
useful to note that Lua is an interpreted language, unlike OCaml, and DWARF
unwind tables are not a concern in Lua.

The interaction on non-local control flow and resources have been studied
extensively. Scheme uses |dynamic-wind| to ensure de-allocation (and
re-allocation) of resources when the non-local control leaves and enters back
into a context. |dynamic-wind| is considered to be complex and not quite what
one wants for resource clean up~\cite{Kiselyov,Sitaram03}. The main challenge
is distinguishing returning exits from a computation closing over resources
from exits that would not return. Resource clean up needs to occur only during
non-returning exit. In Multicore OCaml, we assume that continuations are used
linearly, and hence, no clean up occurs when effects are performed but only
when an exception is thrown, perhaps from the use of |discontinue|. This
obviates the need for |dynamic-wind|.

OCaml does not have a |try/finally| construct available in many programming
languages. In OCaml, alternative standard libraries such as Base~\cite{} and
Core~\cite{} provide mechanisms analogous to Common Lisp |unwind-protect|,
which are in turn implemented using OCaml exceptions. Leijen~\cite{Leijen18}
explicitly extends effect handlers with |finally| clause in Koka. Instead,
Multicore OCaml |discontinue| primitive builds on extant defensive coding
practices against exceptions to clean up resources. This design choice is
crucial to ensure backwards compatibility of legacy systems code with non-local
control flow introduced by effect handlers. Sivaramakrishnan et
al.~\cite{TFP17} describe the interaction of effect handlers and asynchronous
exceptions. This is orthogonal to the contributions of this paper which
focusses on compiler and runtime system support for implementing effect
handlers. Ahman et al.~\cite{Ahman20} present a effect handler core calculus
with asynchrony which might help lay the formal foundations for reconciling
asynchrony in systems and effect handlers.

>>>>

\newcommand{\rot}[2][70]{\adjustbox{angle=#1}{\textbf{#2}}}

@@

Our work is also timely. WebAssembly~\cite{Hass17} community group is actively
considering effect handlers as one of the mechanism for supporting concurrency,
asynchronous I/O and generators~\cite{WasmPropoal}. Project Loom~\cite{loom} is an
OpenJDK project which adds virtual threads and delimited continuations to Java.
Swift concurrency roadmap~\cite{swift} includes direct-style asynchronous
programming and structured concurrency as milestones. We believe that our
design choices will inform similar choices to be made in the other
industrial-strength languages.

@@

\begin{itemize}
  \item the design and implementation of one-shot effect handlers for OCaml.
    Our design retains OCaml's compatibility with program analysis tools that
    inspect the stack using DWARF unwind tables. We have validated our DWARF
    unwind tables using an automated validator tool~\cite{Bastian19}.
  \item a formal operational semantics for the effect handler implementation in
    OCaml.
	\item extensive evaluation which shows that our implementation has negligible
		impact on code that does not use effect handlers, and serves as an
		efficient foundation for scalable concurrent programming.
\end{itemize}

@@

The main challenge in implementing effect handlers in Multicore OCaml is
managing the program stack. In this section, we provide an overview of the
program stack and related mechanisms in stock OCaml.

@@

OCaml does not use the C calling convention. In particular, there are no
callee-saved registers in OCaml. On the |x86-64| backend, the C callee-saved
registers  |r15| holds the \emph{allocation pointer} into the minor heap used
for bump pointer allocation, and |r14| holds a reference to the |Caml_state|
table of global variables used by the runtime. This makes external calls
extremely fast in OCaml. If the external function does not allocate in the
OCaml heap, then the external functions can be called directly, and no
bookkeeping is necessary. For external functions which allocate in the OCaml
heap, the cached allocation pointer is saved to a field in the |Caml_state|
before the external call, and loaded back on return. Similarly, callbacks into
OCaml are also cheap. Callbacks into OCaml involves loading the arguments in
the right registers and calling the OCaml function. OCaml callbacks are
relatively common as the garbage collector (GC) finalisation functions are
executed as OCaml callbacks.

@@

\begin{figure*}
  \begin{minipage}[t]{0.49\linewidth}
  \begin{smathpar}
  \begin{array}{rrcl}
    \text{Constants} & n & \coloneqq & ℤ \\
    \text{Abstractions} & \Lambda & \coloneqq & \lambda^o \mid \lambda^c \\
    \text{Expressions} & e & \coloneqq & n \mid x \mid e ~e \mid  \lam{x}{e} \mid e \odot e \mid \throw{l}{e} \\
                       &   & \mid      & \handle{e}{h} \mid \perform{l}{e} \\
    \text{Handlers} & h & \coloneqq & \{\caseval{x}{e}\} \mid \{\caseexn{l}{x}{e}\} \uplus h \\
                   &   & \mid & \{\caseeff{l}{x}{k}{e}\} \uplus h \\
    \text{Values} & v & \coloneqq & n \mid k \mid \clos{x}{e}{\env} \mid \effval{l}{k} \mid \exnval{l} \\
    \text{Frames} & r & \coloneqq & \farg{e}{\env} \mid \ffun{v} \mid \faritha{\odot}{e}{\env} \mid \farithb{\odot}{ℕ} \\
    \text{Environments} & \env & \coloneqq & \emptyset \mid \envext{\env}{x}{v} \\
  \end{array}
  \end{smathpar}
  \end{minipage}
  \begin{minipage}[t]{0.49\linewidth}
  \begin{smathpar}
  \begin{array}{rrcl}
    \text{Handler Closures} & \hc & \coloneqq & (h,\env) \\
    \text{Frame List} & \fl & \coloneqq & [] \mid r :: \fl \\
    \text{Fibers} & \fiber & \coloneqq & (\fl, \hc) \\
    \text{Continuations} & k & \coloneqq & [] \mid \fiber \kcons k \\
    \text{C stacks} & \cstack & \coloneqq & \cstacka{\fl}{\ostack}\\
    \text{OCaml stacks} & \ostack & \coloneqq & \ostacka{k}{\cstack} \mid \ostackemp \\
    \text{Stacks} & \stack & \coloneqq & \cstack \mid \ostack \\
    \text{Terms} & \term & \coloneqq & e \mid v \\
    \text{Configurations} & \config & \coloneqq & \configa{\tau}{\env}{\stack}
  \end{array}
  \end{smathpar}
  \end{minipage}
	\vspace{-3mm}
  \caption{Syntax of expressions and configurations}
  \label{sem:syntax}
	\vspace{-3mm}
\end{figure*}
\begin{figure*}
	\begin{smathpar}
		\begin{array}{cc}
			\textsc{StepC} ~
			\inferrule{(\term, \env, \fl, \ostack) \cstep \config}
								{\configa{\term}{\env}{\cstacka{\fl}{\ostack}} \step \config} &
			\textsc{StepO} ~
			\inferrule{(\term, \env, k, \cstack) \ostep \config}
								{\configa{\term}{\env}{\ostacka{k}{\cstack}} \step \config}
		\end{array}
	\end{smathpar}
	\mycaption{Top-level reductions}
	\label{sem:toplevel}
	\vspace{-3mm}
\end{figure*}
\begin{figure*}
  \begin{smathpar}
    \begin{array}{rrcl}
      \textsc{Var}     & (x,\env, \fl) & \rightsquigarrow
                       & (\env(x), \env, \fl) \\
      \textsc{Arith1}  & (e_1 \odot e_2, \env, \fl) & \rightsquigarrow
                       & (e_1,\env, \faritha{\odot}{e_2}{\env}::\fl) \\
      \textsc{Arith2}  & (n_1, \_, \faritha{\odot}{e_2}{\env}::\fl) & \rightsquigarrow
                       & (e_2, \env, \farithb{\odot}{n_1}::\fl) \\
      \textsc{Arith3}  & (n_2, \env, \farithb{\odot}{n_1}::\fl) & \rightsquigarrow
                       & (\llbracket n_1 \odot n_2 \rrbracket, \env, \fl) \\
      \textsc{App1}    & (e_1 ~e_2, \env, \fl) & \rightsquigarrow
                       & (e_1, \env, \farg{e_2}{\env}::\fl) \\
      \textsc{App2}    & (\lam{x}{e}, \env, \fl) & \rightsquigarrow
                       & (\clos{x}{e}{\env}, \env, \fl) \\
      \textsc{App3}    & (\clos{x}{e_1}{\env_1}, \_, \farg{e_2}{\env_2}::\fl) & \rightsquigarrow
                       & (e_2, \env_2, \ffun{\clos{x}{e_1}{\env_1}}::\fl) \\
      \textsc{Resume1} & (k,\_,\farg{e_1}{\env_1}::\farg{e_2}{\env_2}::\fl) & \rightsquigarrow
                       & (e_1, \env_1, \ffun{k}::\farg{e_2}{\env_2}::\fl) \\
      \textsc{Resume2} & (\clos{x}{e_1}{\env_1}, \_, \ffun{k}::\farg{e_2}{\env_2}::\fl) & \rightsquigarrow
                       & (e_2, \env_2, \ffun{k}::\ffun{\clos{x}{e_1}{\env_1}}::\fl) \\
      \textsc{Perform} & (\perform{l}{e}, \env, \fl) & \rightsquigarrow
                       & (e, \env, \ffun{\effval{l}{[[],(\{\caseval{x}{x}\},\emptyset)]}}::\fl) \\
      \textsc{Raise}   & (\throw{l}{e}, \env, \fl) & \rightsquigarrow
                       & (e, \env, \ffun{\exnval{l}}::\fl)
    \end{array}
  \end{smathpar}
  \mycaption{Administrative Reductions -- $(\term, \env, \fl) \rightsquigarrow (\tau, \env, \fl)$.}
  \label{sem:step}
	\vspace{-3mm}
\end{figure*}
\begin{figure*}
  \begin{smathpar}
    \begin{array}{rrcll}
      \textsc{AdminC}   & (\term,\env,\fl,\ostack) & \cstep
                        & \configa{\term'}{\env'}{\cstacka{\fl'}{\ostack}}
                        & \text{if } (\term,\env,\fl) \rightsquigarrow (\term',\env',\fl') \\
      \textsc{RetToO}   & (v,\env,[],\ostacka{k}{\cstack}) & \cstep
                        & \configa{v}{\env}{\ostacka{k}{\cstack}} \\
      \textsc{CallC}    & (v, \_, \cclos{x}{e}{\env}::\fl, \ostack) & \cstep
                        & \configa{e}{\envext{\env}{x}{v}}{\cstacka{\fl}{\ostack}} \\
      \textsc{Callback} & (v, \_, \oclos{x}{e}{\env}::\fl, \ostack) & \cstep
                        & \configa{e}{\envext{\env}{x}{v}}{\ostacka{k}{\cstacka{\fl}{\ostack}}}
                        & \text{if } k = [[],(\{\caseval{x}{x}\},\emptyset)] \\
      \textsc{ExnFwdO}  & (v, \env, \ffun{\exnval{l}}::\_, \ostacka{(\fl,\hc) \kcons k}{\cstack}) & \cstep
                        & \configa{v}{\env}{\ostacka{(\ffun{\exnval{l}}::\fl,\hc) \kcons k}{\cstack}}
    \end{array}
  \end{smathpar}
  \mycaption{C Reductions -- $(\term, \env, \fl, \ostack) \cstep \config$.}
  \label{sem:cstep}
	\vspace{-3mm}
\end{figure*}

@@

The operational semantics is an extension of a CEK machine
semantics~\cite{Felleisen86} for effect handlers, following the abstract
machine semantics of Hillerstrom et al.~\cite{Hillerstrom20}. The key
difference from Hillerstrom et al. is that our stacks are composed of
alternating sequence of OCaml and C stack segments. The program state is
captured as configuration $\config \coloneqq \configa{\tau}{\env}{\stack}$ with
the current term $\tau$ under evaluation, its environment $\env$ and the
current stack $\stack$. The term is either an expression $e$ or a value $v$.
The value are integer constants $n$, continuations $k$, closure
$\clos{x}{e}{\env}$, effect being performed $\effval{l}{k}$ and exception being
raised $\exnval{l}$. The environment is a map from variables to values.

@@

The initial configuration for an expression $e$ is
$\configa{e}{\emptyset}{\cstacka{[]}{\bullet}}$, where the environment and the
stack are empty. The top-level reductions (Figure ~\ref{sem:toplevel}) can be
performed by either by taking a C step $\cstep$ or an OCaml step $\ostep$.
Intuitively, this splits the world into executing C code and OCaml code.

@@

\vspace{-2mm}
\subsection{Heap-allocated fibers}

@@

When we first enter OCaml from C, a new fiber is allocated for the main OCaml
stack. Since callbacks may be frequent in OCaml programs that use finalisers,
we run the callbacks on the same fiber as the current one. For example, the
layout of the Multicore OCaml stack at |caml_raise_exn| in the |meander|
example from \S\ref{sec:stack} is shown in Figure~\ref{fig:mcstack}. The
functions |caml_call_c| and |caml_call_ocaml| switch the stacks, and hence are
shown in both the system stack and the fiber. Since we are reusing the fiber
for the callback, care must be taken to save and restore the |handler_info|
before calling and after returning from |c_to_ocaml| function, respectively.
Thanks to the fiber representation, external calls and callbacks remain
competitive with stock OCaml.

\vspace{-2mm}
\subsection{Effect handlers}
\label{sec:effimpl}

@@

\begin{table}
\caption{Micro benchmarks with handlers but no perform. Each entry is the
	slowdown factor ($\times$ times) over its idiomatic implementation in stock OCaml.}
\vspace{-3mm}
{
\begin{tabular}{r c c c c c}
	& \textbf{ack} & \textbf{fib} & \textbf{motzkin} & \textbf{sudan} & \textbf{tak} \\ \hline
	\texttt{MC} 	 	& 10.57 & 15.76 & 16.96 & 12.88 & 13.64 \\
	\texttt{monad} 	& 42.79 & 69.77 & 348.69 & 33.29 & 39.24 \\ \hline
\end{tabular}
}
\label{tab:micro_noperform}
	\vspace{-4mm}
\end{table}

@@

Kawahara et al.~\cite{Kawahara20} implement one-shot effect handlers using
coroutines as a macro-expressible translation, and present an embedding in Lua
and Ruby. Lua provides asymmetric coroutines~\cite{Lua} where each coroutine
uses its own stack similar to how each handled computation runs in its own
fiber in Multicore OCaml. Similar to Multicore OCaml, Lua does not allow
coroutines to include C frames. It is useful to note that Lua is an interpreted
language, unlike OCaml, and DWARF unwind tables are not a concern in Lua.

The interaction on non-local control flow and resources have been studied
extensively. Scheme uses |dynamic-wind| to ensure de-allocation (and
re-allocation) of resources when the non-local control leaves and enters back
into a context. |dynamic-wind| is considered to be complex and not quite what
one wants for resource clean up~\cite{Kiselyov,Sitaram03}. The main challenge
is distinguishing returning exits from a computation closing over resources
from non-returning ones. Resource clean up needs to occur only during
non-returning exit. In Multicore OCaml, we assume that continuations are used
linearly, and hence, no clean up occurs when effects are performed but only
when an exception is thrown, perhaps from the use of |discontinue|. This
obviates the need for |dynamic-wind|.

OCaml does not have a \texttt{\footnotesize try/finally} construct available in
many programming languages. In OCaml, alternative standard libraries such as
Base~\cite{} and Core~\cite{} provide mechanisms analogous to Common Lisp
|unwind-protect|, which are in turn implemented using OCaml exceptions.
Leijen~\cite{Leijen18} explicitly extends effect handlers with |finally| clause
in Koka. Instead, Multicore OCaml |discontinue| primitive builds on extant
defensive coding practices against exceptions to clean up resources. This
design choice is crucial to ensure backwards compatibility of legacy systems
code with non-local control flow introduced by effect handlers.
Sivaramakrishnan et al.~\cite{TFP17} describe the interaction of effect
handlers and asynchronous exceptions. This is orthogonal to the contributions
of this paper which focusses on compiler and runtime system support for
implementing effect handlers. Ahman et al.~\cite{Ahman20} present a effect
handler core calculus with asynchrony which might help lay the formal
foundations for reconciling asynchrony in systems and effect handlers.
